Astr 513: Final Project

### 2016-05-09 Final_Project/ms.tex
Made some grammatical and reference changes
Made some grammar corrections and put in some of the references that
were missing. I’ve also added a few comments where things seem
confusing or an equation might need to be looked at.

\section{Introduction}
\label{sec:Intro}
Type \rom{1}a supernovae (\snea{}) have been instrumental in the discovery 
of the accelerating expansion of the universe. Presently, {\sn} surveys continue 
to be pivotal in constraining cosmological parameters. In this paper, we present 
our analysis of comparing and combining {\sna} inference of cosmological parameters 
to those from other cosmological probes. In particular, we utilize the constraints 
from the Cosmic Microwave Background (CMB) and the Baryonic Acoustic Oscillations 
(BAO) that are gathered in \citet{planck2013}. We look at two cosmologies in our 
analysis: the $\Lambda \textnormal{CDM}$ universe and the $w$CDM universe. This 
paper is outlined as follows. In Section 2, the process and importance of {\snea} 
are summarized. A short description of the \Planck mission is given in Section 3. 
Section 4 describe the two {\sna} surveys whose data we use for our analysis. The 
significance of using distance modulus data to constrain cosmological parameters 
is illustrated in Section 5. The CMB and BAO priors taken from \citet{planck2013} 
are also specified in Section 5. Our results and analysis on the benefit of omitting 
or including Bayesian priors from other cosmological probes 
are in Section 6 and Section 7.

\section{Type \rom{1}a Supernovae}
\label{sec:SNIa}
A type \rom{1}a supernova is a supernova with very little hydrogen or
helium absorption. It is believed that the progenitor for these special
supernovae are white dwarfs that are part of a binary system. These white
dwarfs accrete mass from their companion stars or merge with a companion white
dwarf or neutron star, causing them to exceed the Chandrasekhar limit
($1.44 M_\odot$). When this happens, carbon and oxygen fusion begins in the
core, and the temperature increases. However, the white dwarf does not expand
and cool as a main sequence star would, since it is supported by degeneracy
pressure rather than thermal, so the core experiences a runaway fusion
reaction, releasing $\sim 10^{44}$ J of energy \citep[see, e.g.,][chap. 18.4]{ryden2010}, 
which is enough to overcome the gravitational energy and the star explodes. Because all 
\snea{} are caused by a similar reaction, they all have similar luminosities, and more 
importantly, the light curves all have the same shape: a sharp maximum followed by a smooth, 
gradual decay. Studying nearby \snea{} has shown that the rate of decline of the light curve 
is correlated to the peak luminosity, and this correlation has been calibrated using Cephied 
variable stars located in the same galaxies as nearby \sna{}. With this information, one can 
estimate peak luminosity of an \sna{} from its light curve, thereby providing a 
standard candle with which to measure distance. They were first used in this way in the 
1990s by two independent teams, the Supernova Cosmology Project and the Cal\'{a}n/Tololo 
Supernova Survey. It was using distances from \snea{} from these two projects that 
\citet{riess1998} and \citet{perlmutter1999} were able to discover the accelerating expansion 
rate of the Universe in 1998.

\section{Planck 2015 Results}
\label{sec:planck}
In 2009 the European Space Agency launched the \Planck satellite in order
to measure the Cosmis Microwave Background with unparalleled precision.
By measuring the small anisotropies in the temperature of the CMB across
the entire sky one calculates the angular power spectrum, which can be
modeled using first principles (see, e.g., \citet{dodelson2003}).
By using a likelihood method similar to our own, the authors of
\citet{planck2013} estimated 14 nuisance parameters and 
six cosmological parameters, including the geometrical parameters,
$\Omega_M$, $\Omega_\Lambda$ and $\Omega_k$, that we use in our
{\sna} analysis. The only prior used by \citet{planck2013}
on cosmological parameters was from the WMAP result \citep{wmap2011}
for the optical depth $\tau$. We do not use this parameter
in our {\sna} model.\citet{planck2013} report marginalized
posterior distributions for all parameters,
which we report in Table~XXX.

@@

\subsection{The JLA sample}
\label{sec:betoule}
\citet{betoule2014} define their ``joint light-curve analysis'' (JLA) 
sample, a total of 740 \snea{} compiled from a combination of 
data from the Sloan Digital Sky 
Survey (SDSS), the Supernova Legacy Survey (SNLS), the Hubble Space 
Telescope (HST), and several other experiments. The 613 SDSS \snea{} 
come from the SDSS supernova survey (SDSS-SNS) of SDSS-II \citep{sako2014} 
with redshifts of $0.05 < z < 0.4$, and 374 of them are selected from those 
that were confirmed in the spectroscopic follow-up. The rest of the 
\sne{} come from the compilation by \citet{conley2011}, which includes 
242 spectroscopically confirmed \snea{} from SNLS at redshifts $0.2 < z < 1$, 
14 \snea{} from HST at very high redshifts of $0.7 < z < 1.4$, and 123 
low redshift ($z < 0.08$) from a variety of sources. We made use of the 
publicly available distance moduli, redshifts, and covariance 
matrix provided by \citet{betoule2014}\footnote{Data is available 
for download via 
\url{http://cdsarc.u-strasbg.fr/viz-bin/qcat?J/A+A/568/A22}}. 
We note that the distance moduli are binned, so that they are only computed and 
given on a grid of redshifts. The distance modulus can then be well 
approximated by a piecewise linear function of $\log(z)$ that is given 
in a single redshift bin $z_b \le z < z_{b+1}$ as 
\citep[see][Appendix E.1]{betoule2014}
%
\begin{equation}
\label{eq:binnedMu}
\bar{\mu}(z) = (1 - \alpha) \mu_b + \alpha \mu_{b+1},
\end{equation}
%
where they have defined 
$\alpha = \log\left(\frac{z}{z_b}\right)/\log\left(\frac{z_{b+1}}{z_b}\right)$ 
and $\mu_i$ is the distance modulus at control point $z_i$. 
Therefore, the data and covariance matrix are merely given at the 
31 control points log-spaced in the redshift range $0.01 < z < 1.3$.

\subsection{The Pan-STARRS1 sample}
\label{sec:rest}
\citet{rest2014} uses the first 1.5 year of the first telescope of 
the Panoramic Survey Telescope and Rapid Response System (Pan-STARRS1) 
and a compilation of low-redshift {\sna} studies. The low-redshift 
sample consists of {\snea} collected by Calan/Tololo (Hamuy et al. 1996a), 
CfA1 (Riess et al. 1999), CfA2 (Jha et al. 2006), CfA3 (Hicken et al. 2009a), 
CSP (Contreras et al. 2010), and CfA4 (Hicken et al. 2012). After 
three quality cuts, 113 spectroscopically confirmed {\snea} from 
Pan-STARRS1 ($0.03 < z < 0.65$) and 222 light curves from 197 
low-redshift {\snea} are used for cosmological analysis.

The publicly available data from \citet{rest2014} contains all light 
curves, light curve fit information, covariance matrix, and each {\sn} 
host information.\footnote{Data can be accessed at 
\url{http://ipp0012.ifa.hawaii.edu/SN1aLC/}}. For our analysis, we 
utilized the redshifts, distance moduli, and covariance matrix of the 
final sample of 310 {\snea} or 335 light curves.

@@

\section{Distance Modulus Model}
\label{sec:model}
The main objective of \citet{betoule2014} and \citet{rest2014} is 
to place constraints on cosmological parameters $\Omega_M$, $\Omega_\Lambda$, 
$\Omega_k$, and $w$. In order to so, we must examine the relationship between distance 
modulus and cosmology. The distance modulus $\mu$ is determined from 
the luminosity distance $d_L$ by 
%
\begin{equation}
  \label{eq:mu_dL}
  \mu=5 \log\left(\frac{d_L}{1 \mathrm{Mpc}}\right)+25. 
\end{equation}
%
The luminosity distance depends on the cosmology by 
%
\begin{equation}
\label{eq:luminosityDistance}
d_L(z; w, \Omega_M, \Omega_\Lambda, H_0)= (1+z)|\Omega_k|^{-1/2}\mathcal{S}_{k}\displaystyle\left[c|\Omega_k|^{1/2}\int_{0}^{z}\frac{dz'}{H(z')}\right],
\end{equation}
%
where $\Omega_k \equiv 1-\Omega_M-\Omega_{\Lambda}$. The function 
$\mathcal{S}_{k}$ depends on the curvature density $\Omega_k$ in that 
\[
    \mathcal{S}_{k}= 
\begin{cases}
    \text{sin}(x),&  \Omega_k < 0\\
    x,& \Omega_k = 0\\
    \text{sinh}(x),& \Omega_k > 0
\end{cases}.
\]
The function $H(z)$ is expressed as 
%
\begin{equation}
  \label{eq:hubbleFunction}
  H(z)= H_0\displaystyle\left[\Omega_M(1+z)^3+\Omega_{\Lambda}(1+z)^{3(1+w)}+\Omega_k(1+z)^2\right]^{1/2},
\end{equation}
%
where the Hubble constant $H_0$ can be written in terms of a 
dimensionless Hubble parameter $h$ by $H_0=100h \mathrm{km\ s^{-1}Mpc^{-1}}$.

Two cosmologies are of interest to us: the  $\Lambda$CDM 
universe and the $w$CDM universe. In the  $\Lambda$CDM 
universe, $w=-1$ and constraints are placed on $\Omega_M$ and $h$. 
The $w$CDM universe is a one-parameter extension of  $\Lambda$CDM 
in which $w$ is an arbitrary parameter for a constant dark energy 
equation of state. In $w$CDM, constraints are placed on $\Omega_M$, $w$, 
and $h$. In both cosmologies, the universe is spatially flat 
($\Omega_k = 0$) and thus $\Omega_{\Lambda}=1-\Omega_M$.

\subsection{Likelihood}
\label{sec:likelihood}
Using the above model for the distance modulus allows us to write a
probability that a model can generate our data:
%
\begin{align}
  \label{eq:likelihood}
  P(\mu(z)|\mathrm{cosmo}) &= \frac{1}{\sqrt{2\pi|\mathbf{C}|}}\nonumber \\
  &\qquad{}\exp\left[-\frac{1}{2} \sum_{i,j=0}^{N_{samples}}\left(\mu(z) - \mu_{\mathrm{model}}(z)\right)_i 
    \mathbf{C}^{-1}_{i,j}\left(\mu(z) - \mu_{\mathrm{model}}(z)\right)_j\right]
\end{align}
%
where $\mathbf{C}$ is the full covariance matrix between all of the $N_{samples}$ 
in the data, and $\mathrm{cosmo}$ represents the cosmology that we are investigating
(e.g. flat $\Lambda$CDM, $w$CDM, etc.). 
This is not the full picture though, since we must employ
Bayes' Theorum in order to include any priors on our choice of 
parameters and recover the posterior distribution function of our
parameters given our data
%
\begin{equation}
  \label{eq:posterior}
  P(\mathrm{cosmo}|\mu(z)) = P_{prior}(\mathrm{cosmo}) P(\mu(z)|\mathrm{cosmo}).
\end{equation}
%
For this work we investigate a combination of priors. \CosmoSIS is implemented
in a way that requires the user to provide bounds on the parameter space, which we
detail for each parameter in \autoref{tab:priors}.
Within those bounds, we investigated using a flat prior $P_{prior}=1$ and
\Planck priors.

@@

\section{Results}
\label{sec:results}
The most likely values and uncertainties of $\Omega_M$ and $h$ 
in a $\Lambda$CDM model are listed in \autoref{tab:lcdm}. Included are results from
using the \citet{betoule2014} and \citet{rest2014} data with and without the addition 
of \Planck priors. Additionally, the 68\% and 95\% confidence contours can be seen
in \autoref{fig:betoule_lcdm_corner} and (Rest's corner plot if available).

@@

    \Planck priors. {\it Right}: with \Planck priors.
  \TM{The figure on the right didn't appear in the repo. Please upload.}}

>>>>

\section{Introduction}
\label{sec:Intro}
Type \rom{1}a supernovae (\snea{}) have been instrumental in the discovery 
of the accelerating expansion of the universe. Presently, {\sn} surveys continue 
to be pivotal in constraining cosmological parameters. In this paper, we present 
our analysis of comparing and combining {\sna} inferences of cosmological parameters 
to those from other cosmological probes. In particular, we utilize the constraints 
from the Cosmic Microwave Background (CMB) and Baryonic Acoustic Oscillations 
(BAO) that are gathered in \citet{planck2013}. We look at two cosmologies in our 
analysis: the $\Lambda \textnormal{CDM}$ universe and the $w$CDM universe. This 
paper is outlined as follows. In Section 2, the process and importance of {\snea} 
are summarized. A short description of the \Planck mission is given in Section 3. 
Section 4 describe the two {\sna} surveys whose data we use for our analysis. The 
significance of using distance modulus data to constrain cosmological parameters 
is illustrated in Section 5. The CMB and BAO priors taken from \citet{planck2013} 
are also specified in Section 5. Our results and analysis on the benefit of omitting 
or including Bayesian priors from other cosmological probes 
are in Section 6 and Section 7.

\section{Type \rom{1}a Supernovae}
\label{sec:SNIa}
A type \rom{1}a supernova is a supernova with very little hydrogen or
helium absorption. It is believed that the progenitor for these special
supernovae are white dwarfs that are part of a binary system. These white
dwarfs accrete mass from their companion stars or merge with a companion white
dwarf or neutron star, causing them to exceed the Chandrasekhar limit
($1.44 M_\odot$). When this happens, carbon and oxygen fusion begins in the
core, and the temperature increases. However, the white dwarf does not expand
and cool as a main sequence star would, since it is supported by degeneracy
pressure rather than thermal, so the core experiences a runaway fusion
reaction, releasing $\sim 10^{44}$ J of energy \citep[see, e.g.,][chap. 18.4]{ryden2010}, 
which is enough to overcome the gravitational energy and the star explodes. Because all 
\snea{} are caused by a similar reaction, they all have similar luminosities, and more 
importantly, the light curves all have the same shape: a sharp maximum followed by a smooth, 
gradual decay. Studying nearby \snea{} has shown that the rate of decline of the light curve 
is correlated to the peak luminosity, and this correlation has been calibrated using Cephied 
variable stars located in the same galaxies as nearby \snea{}. With this information, one can 
estimate peak luminosity of a \sna{} from its light curve, thereby providing a 
standard candle with which to measure distance. They were first used in this way in the 
1990s by two independent teams, the Supernova Cosmology Project and the Cal\'{a}n/Tololo 
Supernova Survey. It was using distances from \snea{} from these two projects that 
\citet{riess1998} and \citet{perlmutter1999} were able to discover the accelerating expansion 
rate of the Universe in 1998.

\section{Planck 2015 Results}
\label{sec:planck}
In 2009 the European Space Agency launched the \Planck satellite in order
to measure the Cosmic Microwave Background with unparalleled precision.
By measuring the small anisotropies in the temperature of the CMB across
the entire sky one calculates the angular power spectrum, which can be
modeled using first principles (see, e.g., \citet{dodelson2003}).
By using a likelihood method similar to our own, the authors of
\citet{planck2013} estimated 14 nuisance parameters and 
six cosmological parameters, including the geometrical parameters,
$\Omega_M$, $\Omega_\Lambda$ and $\Omega_k$, that we use in our
{\sna} analysis. The only prior used by \citet{planck2013}
on cosmological parameters was from the WMAP result \citep{wmap2011}
for the optical depth $\tau$. We do not use this parameter
in our {\sna} model. \citet{planck2013} report marginalized
posterior distributions for all parameters,
which we report in \autoref{tab:priors}.
\EW{I'm assuming this should be the \Planck priors table?}

@@

\subsection{The JLA sample}
\label{sec:betoule}
\citet{betoule2014} define their ``joint light-curve analysis'' (JLA) 
sample, a total of 740 \snea{} compiled from a combination of 
data from the Sloan Digital Sky 
Survey (SDSS), the Supernova Legacy Survey (SNLS), the Hubble Space 
Telescope (HST), and several other experiments. The 613 SDSS \snea{} 
came from the SDSS supernova survey (SDSS-SNS) of SDSS-II \citep{sako2014} 
with redshifts of $0.05 < z < 0.4$, and of these, 374 were selected from confirmed \sna{} in the spectroscopic follow-up. The rest of the 
\sne{} came from the compilation by \citet{conley2011}, which includes 
242 spectroscopically confirmed \snea{} from SNLS at redshifts $0.2 < z < 1$, 
14 \snea{} from HST at very high redshifts of $0.7 < z < 1.4$, and 123 
low redshift ($z < 0.08$) from a variety of sources. We made use of the 
publicly available distance moduli, redshifts, and covariance 
matrix provided by \citet{betoule2014}\footnote{Data is available 
for download via 
\url{http://cdsarc.u-strasbg.fr/viz-bin/qcat?J/A+A/568/A22}}. 
We note that the distance moduli are binned, so that they are only computed and 
given on a grid of redshifts. The distance modulus can then be well 
approximated by a piecewise linear function of $\log(z)$ that is given 
in a single redshift bin $z_b \le z < z_{b+1}$ as 
\citep[see][Appendix E.1]{betoule2014}
%
\begin{equation}
\label{eq:binnedMu}
\bar{\mu}(z) = (1 - \alpha) \mu_b + \alpha \mu_{b+1},
\end{equation}
%
where they have defined 
$\alpha = \log\left(\frac{z}{z_b}\right)/\log\left(\frac{z_{b+1}}{z_b}\right)$ 
and $\mu_i$ is the distance modulus at control point $z_i$. 
Therefore, the data and covariance matrix are merely given at the 
31 control points log-spaced in the redshift range $0.01 < z < 1.3$.

\subsection{The Pan-STARRS1 sample}
\label{sec:rest}
\EW{Should we mention the issue with the covariance matrix in this section or the next? Or not at all?}
\citet{rest2014} uses the first 1.5 year of the first telescope of 
the Panoramic Survey Telescope and Rapid Response System (Pan-STARRS1) 
and a compilation of low-redshift {\sna} studies. The low-redshift 
sample consists of {\snea} collected by Cal\'{a}n/Tololo (Hamuy et al. 1996a), 
CfA1 (Riess et al. 1999), CfA2 (Jha et al. 2006), CfA3 (Hicken et al. 2009a), 
CSP (Contreras et al. 2010), and CfA4 (Hicken et al. 2012). After 
three quality cuts, 113 spectroscopically confirmed {\snea} from 
Pan-STARRS1 ($0.03 < z < 0.65$) and 222 light curves from 197 
low-redshift {\snea} are used for cosmological analysis.

The publicly available data from \citet{rest2014} contains all light 
curves, light curve fit information, covariance matrix, and each {\sn} 
host information\footnote{Data can be accessed at 
\url{http://ipp0012.ifa.hawaii.edu/SN1aLC/}}. For our analysis, we 
utilized the redshifts, distance moduli, and covariance matrix of the 
final sample of 310 {\snea} or 335 light curves.

@@

\section{Distance Modulus Model}
\label{sec:model}
The main objective of \citet{betoule2014} and \citet{rest2014} is 
to place constraints on cosmological parameters $\Omega_M$, $\Omega_\Lambda$, 
$\Omega_k$, and $w$. In order to so, we must examine the relationship between distance 
modulus and cosmology. The distance modulus $\mu$ is determined from 
the luminosity distance $d_L$ by 
%
\begin{equation}
  \label{eq:mu_dL}
  \mu=5 \log\left(\frac{d_L}{1 \mathrm{Mpc}}\right)+25. 
\end{equation}
%
The luminosity distance depends on the cosmology by 
%
\begin{equation}
\label{eq:luminosityDistance}
d_L(z; w, \Omega_M, \Omega_\Lambda, H_0)= (1+z)|\Omega_k|^{-1/2}\mathcal{S}_{k}\displaystyle\left[c|\Omega_k|^{1/2}\int_{0}^{z}\frac{dz'}{H(z')}\right],
\end{equation}
%
where $\Omega_k \equiv 1-\Omega_M-\Omega_{\Lambda}$. The function 
$\mathcal{S}_{k}$ depends on the curvature density $\Omega_k$ in that 
\[
    \mathcal{S}_{k}(x)= 
\begin{cases}
    \text{sin}(x),&  \Omega_k < 0\\
    x,& \Omega_k = 0\\
    \text{sinh}(x),& \Omega_k > 0
\end{cases}.
\]
The function $H(z)$ is expressed as 
%
\begin{equation}
  \label{eq:hubbleFunction}
  H(z)= H_0\displaystyle\left[\Omega_M(1+z)^3+\Omega_{\Lambda}(1+z)^{3(1+w)}+\Omega_k(1+z)^2\right]^{1/2},
\end{equation}
%
where the Hubble constant $H_0$ can be written in terms of a 
dimensionless Hubble parameter $h$ by $H_0=100h \mathrm{km\ s^{-1}Mpc^{-1}}$.

Two cosmologies are of interest to us: the  $\Lambda$CDM 
universe and the $w$CDM universe. In the  $\Lambda$CDM 
universe, $w=-1$ and constraints are placed on $\Omega_M$ and $h$. 
The $w$CDM universe is a one-parameter extension of  $\Lambda$CDM 
in which $w$ is an arbitrary parameter for a constant dark energy 
equation of state. In $w$CDM, constraints are placed on $\Omega_M$, $w$, 
and $h$. In both cosmologies, we assume the universe is spatially flat 
($\Omega_k = 0$) and thus $\Omega_{\Lambda}=1-\Omega_M$. 
\EW{Added \textbf{we assume} to the last sentence because $\Lambda$CDM and $w$CDM do not necessarily assume a spatially flat universe}

\subsection{Likelihood}
\label{sec:likelihood}
Using the above model for the distance modulus allows us to write a
probability that a model can generate our data:
%
\begin{align}
  \label{eq:likelihood}
  P(\mu(z)|\mathrm{cosmo}) &= \frac{1}{\sqrt{2\pi|\mathbf{C}|}}\nonumber \\
  &\qquad{}\exp\left[-\frac{1}{2} \sum_{i,j=0}^{N_{samples}}\left(\mu(z) - \mu_{\mathrm{model}}(z)\right)_i 
    \mathbf{C}^{-1}_{i,j}\left(\mu(z) - \mu_{\mathrm{model}}(z)\right)_j\right]
\end{align}
%
where $\mathbf{C}$ is the full covariance matrix between all of the $N_{samples}$ 
in the data, and $\mathrm{cosmo}$ represents the cosmology that we are investigating
(e.g. flat $\Lambda$CDM, $w$CDM, etc.). 
This is not the full picture though, since we must employ
Bayes' Theorum in order to include any priors on our choice of 
parameters and recover the posterior distribution function of our
parameters given our data
%
\begin{equation}
  \label{eq:posterior}
  P(\mathrm{cosmo}|\mu(z)) = P_{prior}(\mathrm{cosmo}) P(\mu(z)|\mathrm{cosmo}).
\end{equation}
%
\EW{Is this equation missing the normalization constant?}
For this work we investigate a combination of priors. \CosmoSIS is implemented
in a way that requires the user to provide bounds on the parameter space, which we
detail for each parameter in \autoref{tab:priors}.
Within those bounds, we investigated using a flat prior $P_{prior}=1$ and
\Planck priors.
\EW{This last sentence is confusing to me...}

@@

\section{Results}
\label{sec:results}
The most likely values and uncertainties of $\Omega_M$ and $h$ 
in a $\Lambda$CDM model are listed in \autoref{tab:lcdm}. Included are results from
using the \citet{betoule2014} and \citet{rest2014} data with and without the addition 
of \Planck priors. Additionally, the 68\% and 95\% confidence contours can be seen
in \autoref{fig:betoule_lcdm_corner} and \autoref{fig:rest_lcdm_corner}.

@@
